---
title: "PracticalMachineLearning_Proj V2"
output: html_document
---
# Synopsis
Data were collected from sensors that are fixed to specific positions on the bodies of 6 participants. Particpants were asked to perfrom specific activities namely "10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions". The data collected from the sensors will be used in this study to tell whether those activities were performed properly or not. More specifically, the study aims to build a machine learning model that can predict how excellent were the performance of these activities and classify it as A for an excellent performance or B,C,D,and E for different types of erratic performances.

#Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

##Data
The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

##Study objective
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

##Author notes
For full reproducibility, the packages below should be installed first, and then the application will set the random number generator seed to 1235. 

#Study and results
##The problem
Data were collected from sensors that are fixed to specific positions on the bodies of 6 participants. Particpants were asked to perfrom specific activities namely "10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions". The data collected from the sensors will be used in this study to tell whether those activities were performed properly or not. More specifically, the study aims to build a machine learning model that can predict how excellent were the performance of these activities and classify it as A for an excellent performance or B,C,D,and E for different types of erratic performances.

##The model

##Exploratory Analysis
##Loading packages
```{r Packages, results='hide'}
library(caret)
library(ggplot2)
library(dplyr)
library(rattle)
library(rpart)
library(rpart.plot)
```
##Initializations
```{r Initializations}
rm(list=ls())
# setting the seed generator number for rerpoducibility
set.seed(1235)
```
##Loading training and testing data
```{r Data_Loading, cache=TRUE}
# Downloading the training data
#trainingFile.url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
#trainingFile=download.file(url=trainingFile.url, destfile=trainingDestfile)
trainingDestfile=".//pml-training.csv"
trainingDataRaw = read.csv(file=trainingDestfile)

# Downloading the testing data
#testingFile.url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
#testingFile=download.file(url=testingFile.url, destfile=testingDestfile)
testingDestfile=".//pml-testing.csv"
testingDataRaw = read.csv(file=testingDestfile)

rm(trainingDestfile, testingDestfile)

#Total number of variables 160 after this stage
```

##Data Processing
###Data Cleaning
```{r Data_Cleaning, cache=TRUE}
# Cleaning the training set
# deleting columns with NAs
trainingDataClean = trainingDataRaw[,colSums(is.na(trainingDataRaw)) == 0]
# Columns 1-7 are irrelevant to our study
irrelevant_indices = 1:7
relevant_indices = - irrelevant_indices
trainingDataClean = trainingDataClean[,relevant_indices]

# Cleaning the testing set
# deleting columns with NAs
testingDataClean = testingDataRaw[,colSums(is.na(testingDataRaw)) == 0]
# Columns 1-7 are irrelevant to our study
irrelevant_indices = 1:7
relevant_indices = - irrelevant_indices
testingDataClean = testingDataClean[,relevant_indices]
rm(trainingDataRaw, testingDataRaw, irrelevant_indices, relevant_indices)
#Total number of variables after this stage, training set: 86 and for intermediate testing: N/A #Testing set: 53
```

###Data Splitting
```{r Data_Splitting, cache=TRUE}
inTrain = createDataPartition(y=trainingDataClean$classe, p=0.75, list= FALSE)
training = trainingDataClean[inTrain,]
testing = trainingDataClean[-inTrain,]

dim(training); dim(testing)

# Initializing the variables for the data to be used in training
# Data Alternative A
trainingStep = training
# Data Alternative B
trainingStepB = training
rm(training, inTrain)

#Total number of variables after this stage, training set A 86 / training set B 86
#Testing set: 53
```
###Exploratory Data Analysis
```{r}

```
###Removing zero- and near-zero-variance variables
```{r ZeroVarianceVariables, cache=TRUE}
# Removing variables that have zero or near-zero variance
nzv = nearZeroVar(trainingStep, saveMetrics = TRUE)
trainingNZV = trainingStep[, !nzv$zeroVar & !nzv$nzv ]
dim(trainingNZV)
trainingStep = trainingNZV
rm(nzv, trainingNZV)

nzvB = nearZeroVar(trainingStepB, saveMetrics = TRUE)
trainingNZVB = trainingStepB[, !nzvB$zeroVar & !nzvB$nzv ]
dim(trainingNZVB)
trainingStepB = trainingNZVB
rm(nzvB, trainingNZVB)

#Total number of variables after this stage, training set A 53 / training set B 53
#Testing set: 53
```
###Removing highly correlated variables
```{r Removing_Correlated_Vars, cache=TRUE}
# Removing highly correlated variables
# training data set A
# removing the classe variable
trainingTemp = trainingStep[,-53]
descrCor <- cor(trainingTemp)
highlyCorrIndices = findCorrelation(descrCor, cutoff = 0.75)
trainingCor = trainingTemp[,-highlyCorrIndices]
# adding the classe variable back
trainingCor$classe = trainingStep$classe
trainingStep = trainingCor

# training data set B
# removing the classe variable
trainingTempB = trainingStepB[,-53]
descrCorB <- cor(trainingTempB)
highlyCorrIndicesB = findCorrelation(descrCorB, cutoff = 0.75)
trainingCorB = trainingTempB[,-highlyCorrIndicesB]
# adding the classe variable back
trainingCorB$classe = trainingStepB$classe
trainingStepB = trainingCorB

rm(trainingCor, trainingCorB, trainingTemp, trainingTempB, descrCor, descrCorB, highlyCorrIndices, highlyCorrIndicesB)

#Total number of variables after this stage, training set A 33 / training set B 33
#Testing set: 53
```
##Choosing final training data
```{r Final_Training_Data, cache=TRUE}
# Training data A for decision trees with no cross-validation
trainingSemifinal = trainingStep

# Training data B for decision trees with 10-fold cross-validation repeated 2 times
trainingSemifinalB = trainingStepB

# Training data C for random forests with 10-fold cross-validation repeated 2 times
trainingSemifinalC = trainingStepB

rm(trainingStep, trainingStepB)

#Total number of variables after this stage, training set A 33 / training set B 33 / 
#training set C 33
#Testing set: 53
```

##Training
###Training using decision trees with no cross-validation
####Training Process
```{r Training_DT_without_cv, cache=TRUE }
modelFit <- train(classe ~ .,method="rpart",data=trainingSemifinal)
fancyRpartPlot(modelFit$finalModel)
```
####Pedictions
```{r Predictiions_DT_without_cv, cache=TRUE}
predicted = predict(modelFit, newdata=testing)
confusionMatrix(predicted, testing$classe)
```
###Training using decision trees with 10-fold cross-validation repeated 10 times
####Training Process
```{r Training_DT_10-fold_cv_rep_2_times, cache=TRUE}
ctrlB = trainControl(method="repeatedcv", number = 10, repeats = 2)
modelFitB <- train(classe ~., data = trainingSemifinalB, method="rpart", tuneLength = 15, trControl=ctrlB)
fancyRpartPlot(modelFitB$finalModel)
```
####Pedictions
```{r Predictions_DT_10-fold_cv_rep_2_times, cache=TRUE}
predictedB = predict(modelFitB, newdata=testing)
confusionMatrix(predictedB, testing$classe)
```
###Training using random forests with 3-fold cross-validation repeated 2 times
####Training Process
```{r Training_RF_3-fold_cv_rep_2_times, cache=TRUE}
# Too much time for 10-cv with 10 repetitions
#ctrlC = trainControl(method="repeatedcv", number = 10, repeats = 10)
#modelFitC <- train(classe ~., data = trainingSemifinalB, method="rf", tuneLength = 10, trControl=ctrlC, verbose = TRUE, trace = TRUE)

# Moderate acceptable time for 3-cv with 2 repetitions
ctrlC = trainControl(method="repeatedcv", number = 3, repeats = 2)
modelFitC <- train(classe ~., data = trainingSemifinalC, method="rf", tuneLength = 2, trControl=ctrlC)
```
####Predictions
```{r Predictions_RF_3-fold_cv_rep_2_times, cache=TRUE}
predictedC = predict(modelFitC, newdata=testing)
confusionMatrix(predictedC, testing$classe)
```
##Writing answers to files
```{r Writing_the_answers, cache=TRUE}
answers = predict(modelFitC, newdata=testingDataClean)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answers)
```
##Results

###Model Evaluation

###Out of sample error


#Conclusion
